\chapter{Evaluation and Discussion}
\label{ch:evaluation}

This chapter presents the evaluation methodology, results, and discussion. The evaluation addresses all three research questions through a combination of automated metrics, an ablation study, and real-world case analysis.


\section{Evaluation Methodology}

\subsection{Evaluation Dataset}

The evaluation dataset comprises 25 proposals: 20 synthetic proposals (P01--P20) and 5 real-world cases (R01--R05).

The 20 synthetic proposals were constructed to provide controlled coverage across risk levels, stratified as 9 High, 7 Medium, and 4 Low risk. Each proposal was authored as a free-text research description of approximately 200--400 words, specifying the AI technology, application domain, data types, affected populations, and deployment context. Ground truth annotations were assigned by the researcher, specifying the expected risk level, applicable requirements (by ID), impacted Charter rights, and relevant incidents. The stratification ensures that the evaluation covers the full range of risk classifications rather than concentrating on a single category.

The synthetic proposal construction protocol follows established benchmarking methodology. PASTA \cite{pasta2026} evaluates AI outputs against multiple compliance policies simultaneously and validates system scores against a panel of human experts' median judgments, directly analogous to AIEF's multi-framework, expert-reviewed design. InvisibleBench \cite{invisiblebench2025}, though situated in a different AI-safety domain (caregiving-relationship AI), demonstrates more broadly that researcher-constructed synthetic scenarios without real-participant data can support controlled, reproducible safety evaluation. LiveRAG \cite{liverag2025} employs a diverse Q\&A dataset for RAG system evaluation, reinforcing the value of controlled synthetic benchmarking for retrieval-augmented systems specifically. Kapania et al. \cite{kapania2024} examine the ethics of using LLMs within HCI research practice, including as substitutes for human input, informing this dissertation's treatment of synthetic proposals as a substitute for confidential real REAMS applications.

The 5 real-world cases provide external validity:

\begin{table}[htbp]
\centering
\caption{Real-world evaluation cases.}
\label{tab:realworld}
\begin{tabular}{llll}
\toprule
\textbf{ID} & \textbf{System} & \textbf{Domain} & \textbf{Expected Risk} \\
\midrule
R01 & Optum algorithm \cite{optum2019} & Healthcare & High \\
R02 & COMPAS \cite{compas2016} & Criminal justice & High \\
R03 & Amazon Rekognition \cite{rekognition2019} & Facial recognition & High \\
R04 & Facebook ad delivery \cite{facebookads2019} & Advertising & High \\
R05 & Epic Sepsis Model \cite{wong2021sepsis} & Healthcare & High \\
\bottomrule
\end{tabular}
\end{table}

These cases were selected because they are extensively documented in academic literature and public reporting, enabling verification of the system's assessments against established analyses of each system's ethical implications.


\subsection{Evaluation Metrics}

Five metrics were used to assess system performance:

\textbf{Fundamental rights coverage (primary).} The proportion of reference-annotation Charter rights surfaced in the system's assessment, calculated as $|R_{addressed} \cap R_{expected}| / |R_{expected}|$. This recall-style claim --- ``did the system surface the rights an annotator identified'' --- is the research plan's primary metric and addresses RQ3. It is reported against reference annotations with the caveat that those annotations are not an external oracle for a ``correct'' ethics assessment.

\textbf{Risk classification accuracy (secondary).} The proportion of proposals for which the system's predicted risk level (High, Medium, or Low) matches the ground truth annotation. Retained as a secondary metric because a single aggregate label collapses heterogeneous per-dimension risk (\S\ref{sec:shacl}).

\textbf{Retrieval recall.} The proportion of ground truth requirement IDs that appear in the knowledge graph retrieval results for a given proposal, calculated as $|R_{retrieved} \cap R_{ground\_truth}| / |R_{ground\_truth}|$. This metric isolates the knowledge graph's contribution independently of the LLM's behaviour.

\textbf{Generation F1 score.} The harmonic mean of precision and recall computed over the requirement IDs cited in the LLM's generated assessment, compared against the ground truth.

\textbf{Documentation completeness (RQ3).} For each authority template, the fraction of required fields that can be derived from a single harmonised assessment JSON without manual input.

\textbf{Ablation and architecture comparison.} Controlled comparison across LLM-only, KG-only, full RAG, and SHACL-only conditions to isolate trade-offs in coverage, determinism, and traceability (RQ2).


\section{The Governance Landscape (RQ1)}
\label{sec:landscape}

Before evaluating the assessment pipeline, this section reports SPARQL analyses over the populated knowledge graph that answer RQ1: where the four frameworks converge, diverge, and leave documented AI harms without corresponding requirement coverage. All figures below were produced by \texttt{analysis/landscape\_analysis.py} against the production GraphDB repository on 18 July 2026.

\begin{table}[htbp]
\centering
\caption{Requirement coverage per Charter right per framework, with incident evidence.}
\label{tab:coverage}
\input{evaluation/coverage_matrix.tex}
\end{table}

\textbf{Overlap distribution.} Of the 18 Charter rights that carry at least one \texttt{mapsToRight} triple from a governance requirement, eight are covered by all four frameworks (Articles~1, 6, 7, 8, 21, 23, 41, and 47 in the requirement IRI naming), six by three frameworks, three by two, and one --- Article~24 (rights of the child) --- by REAMS alone. Convergence is therefore real but uneven: data protection, non-discrimination, dignity, and remedy are shared priorities, while children's rights remain institutionally siloed.

\textbf{Harm without governance.} After merging two dual-IRI aliases (\texttt{Art7\_Privacy}$\rightarrow$\texttt{Art7\_PrivateLife}, \texttt{Art6\_LibertyAndSecurity}$\rightarrow$\texttt{Art6\_RightToLiberty}) and correcting one mapping omission (ACM005 now maps to Article~17 property / IP ownership), \textbf{three} Charter rights remain impacted by curated incidents yet carry zero requirement mappings --- the headline RQ1 finding. Article~15 (freedom to choose an occupation) is impacted by 11 documented incidents; Article~48 (presumption of innocence) by 5; Article~13 (freedom of the arts and sciences) by 1. Each gap was verified against the extracted requirement corpus of all four frameworks to rule out further mapping omission (log in \texttt{analysis/gap\_verification\_log.md}): frameworks address hiring harm via non-discrimination or working-conditions articles, never Art.~15; criminal-conviction \emph{data} requirements are not presumption-of-innocence obligations; no arts/sciences-freedom language appears in any framework text.

\textbf{Governance without documented harm.} The reverse gap also appears. Five rights carry requirement mappings but zero incident evidence in the curated set: Article~3 (right to integrity of the person, 21 requirements), Article~37 (environmental protection, 6), Article~31 (fair working conditions, 5), Article~20 (equality before the law, 4), and Article~22 (cultural diversity, 3). These are not errors --- precautionary or aspirational requirements can precede documented incidents --- but they show where the frameworks encode obligations that the current incident evidence base cannot yet stress-test.

\textbf{Stress-testing the gap under deliberate augmentation.} A finding resting on thin incident evidence (1 incident for Article~13; 5 for Article~48) invites the objection that the gap is an artefact of an under-populated corpus rather than a real governance blind spot. Three additional, independently verifiable incidents were sourced and added specifically to test this: the Dutch SyRI welfare-fraud system, ruled unlawful by The Hague District Court for flagging citizens as fraud suspects on opaque profiling criteria prior to any finding of wrongdoing (Article~48); the discontinuation of multiple US predictive-policing programmes after audits found the underlying data reproduced historical over-policing (Article~48); and documented bias in AI-text-detection tools against non-native English academic writing, restricting researchers' ability to pursue their work under algorithmically-generated suspicion \cite{liang2023gpt} (Article~13). Re-running the landscape analysis with these three incidents added (73 incidents total, up from 70) leaves all three gaps intact: Article~48 now carries 7 incidents and Article~13 carries 2, both still mapped to zero requirements across all four frameworks. The gap survives deliberate stress-testing with genuinely new, real-world evidence; it is not an artefact of an initially thin incident sample.

\textbf{Requirement redundancy.} A separate SPARQL/embedding analysis (\texttt{analysis/requirement\_clustering.py}) tests whether the 207 requirement statements represent 207 distinct obligations or substantively overlapping restatements across frameworks. Sentence embeddings (MiniLM) of all requirement texts were clustered using complete-linkage agglomerative clustering at a cosine-distance threshold of 0.33, calibrated against the empirical 90th-percentile cross-framework similarity (0.670) rather than chosen post-hoc. The 207 statements compress to 180 distinct clusters (13.0\% redundancy), with 25 pairwise and 1 three-way near-paraphrase groups; 9 of these clusters span two distinct frameworks, one covering every pairwise framework combination in the dataset (e.g.\ REAMS R084 and EU AI Act AI029 both asking, in different words, whether pseudonymised data could still permit re-identification; REAMS R087 and Horizon Europe HE052 both asking about the environmental footprint of compute-intensive research). No cluster spans three or more frameworks at this similarity bar: convergence at the sentence level is real but pairwise, a finer-grained and more conservative complement to the article-level convergence reported above.


\section{Multi-Model Evaluation Results}

Table~\ref{tab:model_comparison} presents the primary evaluation results across two model configurations.

\begin{table}[htbp]
\centering
\caption{Evaluation results by model configuration (20 synthetic proposals). Fundamental rights coverage is the primary metric; risk accuracy is secondary.}
\label{tab:model_comparison}
\begin{tabular}{lccc}
\toprule
\textbf{Metric} & \textbf{Llama 3.1 8B} & \textbf{Llama 3.3 70B} \\
\midrule
Fundamental rights coverage (primary) & --- & 0.773 (n=20) \\
\quad 95\% bootstrap CI & --- & [0.632, 0.894] \\
Risk classification accuracy (secondary) & 65\% (13/20) & 90\% (18/20) \\
Retrieval recall & 0.99 & 0.99 \\
Generation F1 & 0.16 & 0.22 \\
Backend & Ollama (local) & Groq API (cloud) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Fundamental Rights Coverage}
\label{sec:rights-coverage}

Mean fundamental rights coverage on the Llama~3.3~70B assessments is 0.773 (n=20) against \texttt{synthetic\_proposals} reference annotations --- the same ground truth used for risk scoring --- with a 95\% bootstrap CI of $[0.632, 0.894]$. Ten proposals achieve full coverage. A string-normalisation check rules out an artefact (``Article~8'' vs ``Article8''). Of 13 per-right misses, \textbf{84\% are generation-stage} (the right was in \texttt{matched\_rights} but not cited in the assessment) and 15\% are retrieval-stage keyword misses (both on Article~11 for P13/P14). That split is consistent with the generation-F1 gap: the knowledge graph surfaces the right, but the LLM selectively omits it. The plan benchmark of $\geq$0.85 was framed against baseline documentation; against reference annotations the CI upper bound reaches 0.894 while the mean sits just below. Risk-label accuracy (90\% at 70B) is retained as secondary. An exact McNemar test --- the recommended paired comparison for classifiers evaluated on the same items \cite{dietterich1998} --- on the 20 paired 8B/70B classifications gives $p=0.125$ (discordant pairs 1 vs 6), consistent with the directional-trend framing rather than a confirmed capacity effect.


\subsection{Central Finding: Model-Agnostic Retrieval}

The most significant finding is that retrieval recall remains at 0.99 across both model configurations. This result demonstrates that the knowledge graph's contribution to the assessment pipeline is independent of the LLM used for generation. The ontology-driven SPARQL retrieval consistently surfaces the correct requirements, incidents, and rights mappings regardless of downstream model capacity.

This finding supports RQ1 and RQ2: the harmonised ontology effectively represents cross-framework requirements, and the knowledge graph supports automated ethics assessment with near-perfect retrieval performance. It also establishes a clean analytical separation between retrieval quality and generation quality, following the decomposition framework proposed by Gao et al. \cite{gao2023rag}. The knowledge engineering contribution is stable and reusable; generation quality is a function of model capacity that will improve as larger and more capable models become available.

\subsection{Controlled Retrieval Baseline: Ontology-Driven vs.\ Lexical Retrieval}
\label{sec:retrieval-baselines}

A high retrieval-recall figure alone does not isolate the ontology's contribution: a simpler retrieval mechanism might achieve the same recall without any knowledge engineering. To test this, the ontology-driven SPARQL retrieval was compared against two standard lexical baselines --- BM25 and TF-IDF cosine similarity --- ranking the same 207 requirement texts for the same 20 proposals, scored as recall against the expected-requirement annotations (Table~\ref{tab:retrieval-baselines}). Because SPARQL retrieval returns a variable-size set (mean 177 requirements), baselines are reported both at fixed cut-offs and at a cut-off matched to the SPARQL set size per proposal.

\begin{table}[htbp]
\centering
\caption{Retrieval recall: ontology-driven SPARQL vs.\ lexical baselines (20 proposals, 16 July 2026).}
\label{tab:retrieval-baselines}
\begin{tabular}{lccccc}
\toprule
\textbf{Condition} & \textbf{R@10} & \textbf{R@25} & \textbf{R@50} & \textbf{R@100} & \textbf{R@$|$SPARQL$|$} \\
\midrule
SPARQL (ontology-driven) & \multicolumn{4}{c}{---} & \textbf{0.992} \\
BM25 & 0.073 & 0.175 & 0.287 & 0.611 & 0.915 \\
TF-IDF cosine & 0.078 & 0.160 & 0.329 & 0.637 & 0.925 \\
\bottomrule
\end{tabular}
\end{table}

The ontology-driven retrieval outperforms both baselines even at matched set size (0.992 vs.\ 0.915/0.925), and the gap widens dramatically at realistic cut-offs: at $k=100$ --- already half the corpus --- lexical similarity recovers only 61--64\% of applicable requirements, because many requirements apply to a proposal through a \emph{rights-mediated} connection (e.g.\ a biometric-surveillance proposal implicating a consent requirement whose text shares no vocabulary with the proposal) that content similarity cannot see but the keyword$\to$right$\to$requirement path in the knowledge graph traverses directly. This is the controlled evidence that the knowledge engineering, not merely retrieval breadth, carries the recall: a lexical retriever given the same generous budget still misses 7.5--8.5\% of applicable requirements. The honest converse, established in \S\ref{sec:ragas-validation}, is that the SPARQL retrieval buys this recall with very low precision; the comparison here shows the trade is nonetheless favourable, since no examined operating point of either baseline dominates it.


\subsection{Risk Classification Analysis}
\label{sec:risk-classification-analysis}

Risk classification accuracy scales with model capacity: 65\% at 8B parameters and 90\% at 70B parameters. The improvement from 65\% (13/20) to 90\% (18/20) reflects a 5-item difference on a 20-item sample and should be read as a directional trend consistent with the expected relationship between model capacity and reasoning accuracy, rather than a statistically established effect; no significance test was applied given the sample size, and a larger evaluation set is needed before the capacity-accuracy relationship can be treated as confirmed rather than suggestive. Analysis of the 8B model's misclassifications reveals a consistent pattern of conservative bias: the model over-classifies Medium and Low risk proposals as High risk. This bias does not manifest as under-detection of genuine high-risk proposals, which are correctly classified at a 90\% rate even by the 8B model.

In the context of ethics assessment, this conservative behaviour is defensible. False negatives (failing to flag a genuinely risky proposal) carry greater potential for harm than false positives (flagging a low-risk proposal for additional review). This asymmetry aligns with the precautionary principle embedded in EU AI Act Article~9, which requires risk management systems to err on the side of caution. Farber \cite{farber2025} reports an analogous finding in offender risk assessment, where AI models consistently assign higher overall risk scores than human assessors, interpreting this as a characteristic of LLM-based risk assessment rather than an error.

The 90\% accuracy achieved by the 70B model demonstrates that the system produces reliable risk classifications when provided with sufficient model capacity.


\subsection{Generation F1 Analysis}

Generation F1 scores (0.16 at 8B, 0.22 at 70B) are lower than retrieval recall, reflecting the expected gap between retrieving the correct requirements and the LLM's ability to cite them accurately in its generated assessment. The LLM receives all relevant requirements in its context but selectively cites a subset, sometimes omitting requirements that are applicable or occasionally citing requirements that were retrieved but are marginally relevant to the specific proposal.

This finding confirms the analytical value of separating retrieval and generation metrics: the knowledge graph performs its function well (0.99 retrieval recall), while generation quality is a separable concern that improves with model capacity and can be further enhanced through prompt engineering or fine-tuning.


\subsection{Qwen3 32B: Exploratory Results}
\label{sec:qwen-ratelimit}

An exploratory evaluation was conducted using Qwen3 32B via the Groq API. However, 7 of 20 proposals failed to complete due to tokens-per-minute (TPM) rate limiting on the free API tier. The results from the 13 completed proposals are therefore methodologically compromised and cannot be reliably compared against the Llama configurations. For this reason, Qwen3 32B results are excluded from the primary comparison. This limitation is acknowledged as a constraint of the evaluation infrastructure rather than a reflection on the model's capabilities.

\subsection{Error Taxonomy and Per-Framework Analysis}
\label{sec:error-taxonomy}

Every deviation from ground truth across the 20 proposals (Llama 3.3 70B configuration) was classified into a seven-category taxonomy (Table~\ref{tab:error-taxonomy}), separating errors by pipeline stage (retrieval vs.\ generation) and object (risk tier, requirements, rights).

\begin{table}[htbp]
\centering
\caption{Error taxonomy over 20 proposals (Llama 3.3 70B).}
\label{tab:error-taxonomy}
\begin{tabular}{llr}
\toprule
\textbf{Category} & \textbf{Stage} & \textbf{Count} \\
\midrule
E1: Risk over-classification & generation & 2 \\
E2: Risk under-classification & generation & \textbf{0} \\
E3: Requirement retrieval miss & retrieval & 2 \\
E4: Requirement generation miss (retrieved, not cited) & generation & 180 \\
E5: Rights generation miss (matched, not cited) & generation & 11 \\
E6: Rights retrieval miss & retrieval & 2 \\
E7: Citations outside expected set (reported, not scored as error) & generation & 130 \\
\bottomrule
\end{tabular}
\end{table}

Three findings follow. First, the conservative-bias claim of \S\ref{sec:risk-classification-analysis} is now exact: across all 20 proposals the system \emph{never} under-classifies risk (E2 $=0$); its only tier errors are upward (E1 $=2$) --- precisely the error asymmetry the precautionary posture of ethics screening prefers. Second, the pipeline's bottleneck is overwhelmingly generation-stage: of 182 missed requirement citations, 180 (98.9\%) had been successfully retrieved and were present in the prompt context, against only 2 genuine retrieval misses; the corresponding split for rights is 11 vs.\ 2. Third, category E7 is reported but deliberately not scored as error: a citation outside the expected-requirement annotation is frequently a defensible additional finding rather than a mistake, and scoring it as error would presuppose the reference annotations are exhaustive --- which, as a single-annotator artifact, they are not.

\begin{table}[htbp]
\centering
\caption{Per-framework requirement performance (20 proposals, Llama 3.3 70B).}
\label{tab:per-framework}
\begin{tabular}{lrrrrr}
\toprule
\textbf{Framework} & \textbf{Expected} & \textbf{Retrieved} & \textbf{Ret.\ recall} & \textbf{Cited} & \textbf{Cite recall} \\
\midrule
REAMS & 104 & 102 & 0.981 & 7 & 0.067 \\
EU AI Act & 42 & 42 & 1.000 & 10 & 0.238 \\
Horizon Europe & 36 & 36 & 1.000 & 9 & 0.250 \\
ACM/NeurIPS & 37 & 37 & 1.000 & 11 & 0.297 \\
\bottomrule
\end{tabular}
\end{table}

The per-framework breakdown (Table~\ref{tab:per-framework}) exposes a systematic citation skew: retrieval recall is near-uniform across frameworks, but the LLM's citation recall is 4.4$\times$ lower for REAMS (0.067) than for ACM/NeurIPS (0.297). REAMS contributes the most expected requirements per proposal (104 across 20 proposals) while the model cites only $\sim$5 requirements per assessment in total, so the largest framework is proportionally most under-represented in generated citations. This is a concrete, previously invisible consequence of leaving citation selection to the LLM, and a direct argument for the deterministic per-dimension layer of \S\ref{sec:shacl}: rule-based verdicts do not exhibit budget-driven framework skew.


\section{Ablation Study}
\label{sec:ablation}

An ablation study was conducted on five proposals (P01, P03, P07, P13, P17) selected to span all three risk levels. Three conditions were evaluated:

\begin{table}[htbp]
\centering
\caption{Ablation study results.}
\label{tab:ablation}
\begin{tabular}{p{2.5cm}p{4cm}ccr}
\toprule
\textbf{Condition} & \textbf{Description} & \textbf{Req. IDs} & \textbf{Incidents} & \textbf{Rights} \\
\midrule
A: LLM Only & Proposal sent to LLM without KG context & None & 0--1 & 1--2 \\
B: KG Only & SPARQL retrieval without LLM interpretation & All & 8 & 9--11 \\
C: Full Pipeline & KG retrieval + LLM interpretation & Cited & 1--2 & 2--4 \\
\bottomrule
\end{tabular}
\end{table}

The ablation study reveals three concrete contributions of the knowledge graph:

\textbf{Requirement traceability.} Under Condition A (LLM-only), the model produces general ethics commentary but cites no specific requirement IDs from any governance framework. The LLM lacks the structured knowledge to provide framework-specific, traceable compliance guidance. Under Condition C (full pipeline), assessments cite specific requirement IDs with framework provenance (e.g., ``R042 from REAMS \S 3.1.4''), enabling the user to verify each recommendation against the source document.

\textbf{Incident grounding.} Under Condition A, the LLM references 0--1 incidents, which are frequently hallucinated (naming incidents that do not exist or misattributing details). Under Condition B (KG-only), retrieval surfaces 8 relevant incidents per proposal from the curated dataset. Under Condition C, the LLM contextualises 1--2 of the most relevant retrieved incidents in its narrative, grounding its assessment in documented precedent rather than fabricated examples.

\textbf{Rights coverage.} Under Condition A, the LLM identifies 1--2 Charter rights, typically the most obvious (e.g., data protection for any project involving personal data). Under Condition B, the KG retrieves 9--11 mapped rights per proposal, reflecting the full breadth of Charter implications. Under Condition C, the LLM synthesises 2--4 rights in its assessment, providing broader coverage than the LLM-alone baseline while remaining manageable for the reader.

These findings, together with the architecture comparison in Table~\ref{tab:arch_compare}, address RQ2: the hybrid RAG pipeline trades determinism for free-text coverage and narrative explanation, while SHACL (\S\ref{sec:shacl}) trades coverage for full scrutability and editability.

\begin{table}[htbp]
\centering
\caption{Architecture comparison for RQ2 (LLM-only vs hybrid RAG vs SHACL-only).}
\label{tab:arch_compare}
\begin{tabular}{p{2.2cm}p{3.0cm}p{3.0cm}p{3.2cm}}
\toprule
\textbf{Property} & \textbf{LLM-only} & \textbf{Hybrid RAG} & \textbf{SHACL-only} \\
\midrule
Input type & Free text & Free text & Structured TTL (or keyword-bridged text) \\
Determinism & Stochastic & Stochastic generation; deterministic retrieval & Fully deterministic \\
Traceability & No requirement IDs; hallucinated incidents & Requirement IDs + incident IRIs with provenance & Shape IRI, SPARQL pattern, priority triple \\
Per-dimension output & Collapsed into narrative / single label & Narrative + rights list; single risk label & Per-dimension profile ordered by priority \\
Editability & Prompt / model change only & Ontology + prompt; ranking inside LLM & Edit one \texttt{aiefsh:priority} triple and re-validate \\
Rights P / R (n=20) & --- & coverage recall 0.773 & 0.566 / 0.327 \\
New-requirement update cost & Prompt engineering (untested effort) & Add TTL individual + full-graph reload ($\sim$30s reload, plus a full re-evaluation run to confirm no regression) & Append one shape ($\sim$14s syntax check), independently testable, no reload \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Widening the rule set (Week-2 methodological addition).} An initial six-rule, four-dimension SHACL demonstrator invited a fair objection: comparing a narrow rule set against the full-coverage RAG pipeline is not a fair architecture comparison, whatever the caveats. The rule set was therefore widened to 21 shapes covering 20 of the 22 Charter dimensions actually mapped in the knowledge graph, using the same shape pattern (dimension, priority, severity, rationale, SPARQL constraint) throughout. Two dimensions (Art~13 Freedom of Arts and Science; Art~20 Equality Before the Law) remain unruled because no reliable keyword signal in free text distinguishes them from already-ruled adjacent dimensions without a materially richer NLU layer than the keyword bridge supports --- a stated scope limit rather than an omission. Widening moved mean precision from 0.417 to \textbf{0.566} and mean recall from 0.193 to \textbf{0.327}, confirming the earlier recall figure was bounded by rule-set coverage rather than an intrinsic ceiling of the SHACL approach; recall still trails the RAG pipeline's rights coverage (0.773), which is the expected and honest result of comparing 21 deterministic rules against an LLM with broader, if less precise and less traceable, interpretive reach.

\textbf{Time-to-update.} To make the editability claim concrete rather than asserted, one new governance obligation (AI Act Art.~12 record-keeping) was added to both systems under identical conditions, timed, and then reverted (this was a controlled trial to measure process cost, not a permanent ontology extension, so the dissertation's requirement and rule counts elsewhere are unaffected). Adding the KG-side requirement required drafting an 11-line Turtle individual (identifier, text, framework, tier, legal basis, one \texttt{mapsToRight} triple), then clearing and reloading the full production GraphDB repository ($\sim$30 seconds measured for the reload-and-verify step alone) --- and, critically, no automated check confirms the addition actually changes retrieval or generation behaviour without re-running the evaluation pipeline, so silent regressions are possible. Adding the equivalent SHACL rule required an 11-line shape block appended to the existing file, syntactically validated in under a second via \texttt{rdflib}, immediately testable in isolation against any proposal via \texttt{run\_shacl.py} with zero risk of side-effects on the other shapes, and expressed as a single, minimal, human-reviewable git diff. The authoring effort is comparable (both $\sim$11 lines); the operational risk profile is not: the RAG-side change requires a full-corpus reload and an implicit trust that retrieval logic still behaves correctly, while the SHACL-side change is atomic, isolated, and verifiable before it is ever run against real data.

SHACL mean precision 0.566 and recall 0.327 over the 20 synthetic proposals, now spanning 20 of 22 Charter dimensions, are the RQ2 trade-off stated with numbers: still narrower than the hybrid pipeline's rights coverage, but no longer narrow because of an artificially small rule set. A deterministic citation post-filter enforcing $\mathrm{citation}\subseteq\mathrm{retrieval}$ found a 0\% out-of-retrieval citation rate (0/167 after normalising prefixed ID variants such as \texttt{EUAIAct AI001}$\rightarrow$\texttt{AI001}); post-filtering leaves F1 unchanged at 0.227, so grounding works and the transparency panel's divergence warnings rarely fire.


\section{Documentation Generation (RQ3)}
\label{sec:docgen}

Objective~4 of the original research plan --- one assessment producing authority-formatted documentation --- is implemented by \texttt{rag-pipeline/generate\_documentation.py}. A single Llama~3.3~70B assessment JSON is projected onto four templates (REAMS, FRIA, ACM/NeurIPS, Horizon Europe), and completeness is scored as the fraction of required fields filled without manual input.

\begin{table}[htbp]
\centering
\caption{Mean documentation completeness from one harmonised assessment (20 synthetic proposals, Llama~3.3~70B).}
\label{tab:doc_completeness}
\begin{tabular}{lrr}
\toprule
\textbf{Authority} & \textbf{Mean completeness} & \textbf{n} \\
\midrule
FRIA & 100\% & 20 \\
ACM/NeurIPS & 100\% & 20 \\
REAMS & 86\% & 20 \\
Horizon Europe & 61\% & 20 \\
\bottomrule
\end{tabular}
\end{table}

FRIA and ACM/NeurIPS fields map cleanly onto the assessment schema (purpose, rights, risks, mitigations, precedents). Field-level breakdown shows the unfilled slots concentrated in Horizon Europe \texttt{Human Oversight} (19/20 docs), REAMS \texttt{Consent Arrangements} (17/20), and Horizon \texttt{Data Management} (12/20). Consent is a REAMS-specific checklist heading that the assessment only fills when a cited requirement literally mentions consent --- a docgen mapping thinness. Human oversight, by contrast, is demanded by Horizon's Ethics-by-Design guidance in a form no other framework's template requires as a standalone section; the completeness shortfall is therefore partly a measurement of Horizon's structural distinctiveness, not solely a system limitation. Generated Markdown artefacts for all 20 proposals are in \texttt{evaluation/generated-documentation/}.


\section{Real-World Case Analysis}

The five real-world cases (R01--R05) were re-processed through the deployed production pipeline (\url{https://aief.vercel.app}, Llama 3.3 70B backend) on 14 July 2026 to assess external validity, so the figures below are directly reproducible rather than reported from memory. All five cases were correctly classified as High risk. The system's assessments were qualitatively compared against published analyses of each system's ethical implications:

\textbf{R01 (Optum).} The system correctly identified healthcare discrimination risk and mapped the case to Article~21 (non-discrimination) and Article~35 (healthcare), citing requirements ACM023, ACM027, HE032, and R021, and retrieving the Optum healthcare-bias incident itself (AIAAIC-006) alongside a further healthcare-domain precedent (AIAAIC-043).

\textbf{R02 (COMPAS).} The system identified criminal justice bias, mapping to Article~21, Article~6 (liberty and security of person), and Article~47 (right to effective remedy), citing requirement R068 and retrieving the Detroit false-arrest incident (AIAAIC-002) and a further law-enforcement precedent (AIAAIC-038).

\textbf{R03 (Amazon Rekognition).} Facial recognition bias was correctly identified with mappings to Article~7 (private life), Article~8 (data protection), and Article~21, citing requirements ACM024, AI024, and HE048. The system retrieved the Detroit false-arrest incident (AIAAIC-002) as a cross-technology precedent from the same risk category, alongside two further precedents (AIAAIC-007, AIAAIC-051).

\textbf{R04 (Facebook ad delivery).} Discriminatory ad targeting was identified with mappings to Article~21 and Article~23 (equality between women and men), citing requirements ACM003, ACM004, HE015, AI023, and R054, and retrieving the Amazon hiring-bias incident (AIAAIC-001) and a further precedent (AIAAIC-011).

\textbf{R05 (Epic Sepsis Model).} An earlier draft of this dissertation used an unverifiable ``NHS sepsis prediction algorithm'' as R05; that citation has been corrected. R05 is now the Epic Sepsis Model, a proprietary clinical decision-support tool deployed across hundreds of US hospitals, whose external validation at a large academic medical centre found substantially worse real-world discrimination than the developer reported (AUC 0.63 versus a claimed 0.76--0.83), 33\% sensitivity for genuine sepsis cases, a 12\% positive predictive value, and a high false-alarm burden contributing to clinician alert fatigue \cite{wong2021sepsis}. AIEF classified this case as High risk, citing requirements ACM003, HE013, and R085, mapping to Article~21, and retrieving the Optum healthcare-bias incident (AIAAIC-006) and the Amazon hiring-bias incident (AIAAIC-001) as precedents. This revises the earlier draft's characterisation of R05 as Medium risk on the premise that clinical oversight mitigates model error; the peer-reviewed evidence on the Epic Sepsis Model does not support that premise, since alert fatigue from a high false-positive rate is itself a documented mechanism by which human oversight fails.

These results provide evidence that the system generalises beyond the synthetic evaluation dataset to documented real-world cases, though the sample size (n=5) limits the strength of this claim.


\section{RAGAS-Style Automated Metrics}
\label{sec:ragas-validation}

To complement the risk-classification-accuracy, retrieval-recall, and generation-F1 metrics reported above, four RAGAS-style metrics \cite{gao2023rag} were computed for all 20 synthetic proposals on 14 July 2026: context recall and context precision, computed directly against live SPARQL retrieval from the GraphDB endpoint (no LLM judge required), and faithfulness and answer relevancy, scored by Llama 3.3 70B as an LLM judge against the generated assessments.

\textbf{A data-consistency issue was identified and corrected between the 14 and 15 July runs.} The 14 July run queried the production GraphDB repository (\texttt{ai-ethics-kg}) that \texttt{sparql\_retrieval.py} is hard-coded to use. Although the requirement graph had been deduplicated in the source Turtle file on 13 July (removing conflicting duplicate blocks for R010 and R054, \S\ref{sec:kg}), that corrected file had not yet been reloaded into this production repository — a distinct oversight from the file-level fix. Direct verification confirmed the production repository still returned two conflicting \texttt{requirementText} values for R010 as of 14 July. The repository was cleared and reloaded from the corrected Turtle file on 15 July, and all four RAGAS metrics were re-run. Context recall and context precision were verified \emph{bit-for-bit identical} before and after the fix (0.7962326562326563 in both runs to sixteen significant figures), confirming that the duplicate-record issue was namespace- and property-value-scoped and did not corrupt identifier-level retrieval matching; requirement IDs remained consistent across both graph states because the retrieval queries filter by IRI, and the duplicate blocks described the same IRI rather than introducing new ones. Faithfulness and answer relevancy, being independent LLM-judge calls, were re-run in full and gained substantially fuller coverage in the process. The figures below are from the 15 July re-run.

Context recall and context precision completed for all 20 proposals in both runs. The LLM-judge metrics (faithfulness, answer relevancy) completed for 5/20 and 12/20 proposals respectively on 14 July, before repeatedly hitting the Groq free-tier tokens-per-minute rate limit already documented for the Qwen3 32B exploratory evaluation (\S\ref{sec:qwen-ratelimit}); on 15 July's re-run, both completed for 19/20 proposals (P20 was excluded from both by construction, having retrieved zero context).

\begin{table}[htbp]
\centering
\caption{RAGAS-style metrics (20 synthetic proposals, 15 July 2026 re-run).}
\label{tab:ragas}
\begin{tabular}{lrrrr}
\toprule
\textbf{Metric} & \textbf{Mean} & \textbf{Range} & \textbf{n} & \textbf{Judge} \\
\midrule
Context recall & 0.80 & 0.00--1.00 & 20/20 & none (set overlap) \\
Context precision & 0.030 & 0.006--0.065 & 20/20 & none (set overlap) \\
Faithfulness & 0.35 & 0.0--0.8 & 19/20 & Llama 3.3 70B \\
Answer relevancy & 0.88 & 0.5--1.0 & 19/20 & Llama 3.3 70B \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Context recall} (mean 0.80) is lower than the 0.99 headline retrieval-recall figure reported in Table~\ref{tab:model_comparison} because the two metrics use different denominators: the primary retrieval-recall metric checks recall of ground-truth requirement IDs used for the risk-classification evaluation, while RAGAS context recall here is computed against the full 5--13-item expected-requirement sets used for generation-F1 scoring, a stricter test. One proposal (P20, open-source code linting, Low risk) returned zero recall — the single expected requirement (ACM001) was not among the retrieved set, a genuine keyword-retrieval failure on a minimal-risk, technically-phrased proposal that the keyword map does not cover well, and is recorded as a concrete limitation of the current keyword-based retrieval mechanism.

\textbf{Context precision} (mean 0.030) should not be read as a tuned design parameter but as a consequence of keyword-based SPARQL retrieval being deliberately permissive: with 140--196 of the 207 total requirements typically surfaced per proposal, the retrieval stage functions closer to broad candidate filtering than to precise, targeted matching, leaving the LLM to filter down to the requirements it ultimately cites (a mean of roughly 5 per proposal, per the ablation study in \S\ref{sec:ablation}). This was an acceptable trade-off given the ethics-screening context, where a missed requirement (false negative) is more consequential than an irrelevant one being briefly surfaced (false positive). However, it also means the RAG pipeline's contribution is closer to ``near-complete context plus LLM triage'' than to precision-oriented retrieval, and the low precision should be read as a current limitation of the retrieval design rather than an intentionally optimised operating point.

\textbf{Faithfulness} (mean 0.35, n=19) sits well below \textbf{answer relevancy} (mean 0.88, n=19), a gap worth confronting directly rather than reporting in isolation. Three explanations are plausible and not mutually exclusive: first, the judge model (Llama 3.3 70B) is drawn from the same model family as the system under evaluation, which self-consistency literature suggests can bias faithfulness scoring in either direction depending on shared failure modes; second, retrieval surfaces 140--196 requirements per proposal (discussed above), and an LLM asked to synthesise a narrative from a near-complete requirement set may draw on background knowledge to fill gaps rather than citing only what was retrieved, which RAGAS-style faithfulness scoring penalises even when the output is substantively correct; third, faithfulness and answer relevancy are themselves LLM-judged scores and carry their own measurement uncertainty, evidenced directly by the fact that the two runs produced materially different means (0.30 on 14 July's 5-proposal partial sample versus 0.35 on 15 July's 19-proposal re-run) from queries against an unchanged set of generated assessments. The high answer relevancy alongside low faithfulness suggests the pipeline produces on-topic, plausible assessments whose surface content is not always traceable to specific retrieved passages — a distinction that matters for a tool whose central design goal is traceability, and one flagged as a priority for follow-up work in Chapter~\ref{ch:conclusion} rather than a settled negative result.


\section{Discussion}

\subsection{Addressing the Research Questions}

\textbf{RQ1} asked where the four frameworks converge, diverge, and leave documented harms without requirement coverage. Section~\ref{sec:landscape} answers directly: eight rights are covered by all four authorities, Article~24 by REAMS alone, and three verified genuine gaps (Arts.~13, 15, 48) after dual-IRI cleanup and one mapping correction.

\textbf{RQ2} asked about trade-offs between scrutable symbolic rules and LLM-based interpretation. Table~\ref{tab:arch_compare} and \S\ref{sec:shacl} summarise the comparison: hybrid RAG accepts free text and produces narrative assessments with requirement provenance but stochastic ranking; SHACL accepts structured (or keyword-bridged) descriptions and produces deterministic, per-dimension, editable profiles at the cost of narrower coverage. The ablation study shows that LLM-alone lacks traceable requirement IDs and reliable incident grounding.

\textbf{RQ3} asked whether one harmonised assessment can satisfy divergent authority documentation structures, and with what rights coverage against reference annotations. Documentation completeness reaches 100\% for FRIA and ACM/NeurIPS, 86\% for REAMS, and 61\% for Horizon Europe (\S\ref{sec:docgen}). Mean fundamental rights coverage is 0.773 (n=20, 95\% CI $[0.632, 0.894]$) against reference annotations (\S\ref{sec:rights-coverage}), with 84\% of misses at generation stage.


\subsection{Comparison with Related Work}

The conservative bias observed in AIEF's smaller model mirrors findings by Farber \cite{farber2025} in a different risk assessment domain, suggesting this may be a general characteristic of LLM-based risk assessment that warrants further investigation.


\subsection{Threats to Validity}

\textbf{Internal validity.} The ground truth annotations for synthetic proposals were assigned by the researcher rather than an independent panel. While the inter-annotator agreement study (Cohen's $\kappa$ = 0.712) validates the annotation methodology for incident-rights mappings, the risk level assignments for synthetic proposals represent a single annotator's judgment.

\textbf{External validity.} The evaluation uses synthetic proposals because real REAMS applications are confidential. The 5 real-world cases mitigate this limitation but do not fully replicate the conditions under which the system would be used in practice. The system has not been evaluated by REAMS committee members or ethics reviewers.

\textbf{Construct validity.} Risk classification accuracy treats risk levels as categorical labels, but in practice, the boundary between Medium and High risk is often contextual and debatable. The 65\% accuracy at 8B may partly reflect ambiguity in ground truth labels rather than system error.

\textbf{Reliability.} LLM outputs are stochastic. Running the same proposal through the pipeline multiple times may produce slightly different assessments. The evaluation reports results from a single run per proposal. Temperature and other generation parameters were held constant across all evaluations. Given that risk classification accuracy is the dissertation's central quantitative claim, repeated runs per proposal (to characterise output variance) and a second independent annotator for ground-truth labelling are identified as immediate priorities for strengthening this result, rather than generic future work.

\textbf{How this evaluation should be read.} Nearly every quantitative claim in this chapter --- risk-level labels, expected-requirement sets, expected-rights annotations, and the requirement$\to$right mappings underlying the landscape analysis of \S\ref{sec:landscape} --- traces back to annotations produced by the same person who designed and built the system being evaluated. This is stated plainly rather than left for an examiner to discover: every number in this dissertation should be read first and foremost as evidence of \emph{internal consistency} between AIEF and its designer's own annotation scheme, not as evidence of external, independently-verified correctness. Three sources of evidence in this dissertation do not share that annotation origin and should be weighted accordingly when judging external validity: the OOPS! pitfall scan and HermiT reasoner check (\S\ref{sec:oops-validation}, \S\ref{sec:cq-validation}), which are automated and annotation-independent; the Cohen's $\kappa=0.712$ inter-annotator agreement study on incident-rights mappings, which involved a second coder; and the expert review scores solicited from the supervisory team, reported separately where available. Where this dissertation reports a number without one of these three forms of external check, that number should be understood as a measurement of the system against itself.

\subsection{Limitations}

Several limitations should be acknowledged:

\begin{enumerate}
\item The ontology covers four governance frameworks; additional frameworks (e.g., ISO/IEC 42001, NIST AI RMF) are not included and would require ontology extension.

\item The 70 incidents represent a curated subset of available AI incident data. As new incidents occur (362 in 2025 alone \cite{stanfordaiindex2026}), the knowledge graph requires periodic updates.

\item Expert validation of system outputs was planned but not completed within the project timeline. Scoring sheets were prepared for 8 proposals across ADAPT Centre researchers, but responses were not received before submission.

\item The Qwen3 32B evaluation was compromised by API rate limits, preventing a complete three-model comparison.

\item The web application uses a static knowledge graph snapshot for deployment, which does not update automatically as new triples are added to GraphDB.

\item The system's SPARQL retrieval relies on keyword matching, which may miss requirements that are relevant but expressed using different terminology than the proposal text.
\end{enumerate}
