\documentclass[a4paper,oneside,12pt]{book}

%----------------------------------------------------------------------------------------
%	THESIS INFORMATION
%----------------------------------------------------------------------------------------

\newcommand{\thesistitle}{AI-Assisted Ethics Assessment Framework for Early-Stage Research Projects: An Ontology-Driven Knowledge Graph and Retrieval-Augmented Generation Approach}
\newcommand{\degree}{MSc Computer Science (Intelligent Systems)}
\newcommand{\typeofthesis}{dissertation}
\newcommand{\authorname}{Navina Ganapathy Amuthan}
\newcommand{\supervisor}{Prof. Dave Lewis}
\newcommand{\keywords}{AI ethics, ontology, knowledge graph, retrieval-augmented generation, EU AI Act, fundamental rights, ethics assessment, REAMS, compliance automation}
\newcommand{\school}{\href{https://www.tcd.ie/scss/}{School of Computer Science and Statistics}}
\newcommand{\college}{Trinity College Dublin}

%% Language and font encodings
\usepackage[T1]{fontenc} 
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

%% Bibliographical stuff
\usepackage[]{cite}
%% Document size
\usepackage[a4paper,top=2.54cm,bottom=2.54cm,left=2.54cm,right=2.54cm,headheight=16pt]{geometry}
\setlength{\marginparwidth}{2cm}
%% Useful packages
\usepackage{amsmath}
\usepackage[autostyle=true]{csquotes}
\usepackage[pdftex]{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{hyperxmp}
\usepackage{caption}
\usepackage{sfmath}
\usepackage[parfill]{parskip}
\usepackage{setspace}
\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{fancyhdr}
\usepackage{xcolor}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{listings}
\usepackage{url}
\numberwithin{equation}{chapter}
\pagestyle{plain}

\definecolor{tcd_blue}{RGB}{5, 105, 185}

\renewcommand{\familydefault}{\sfdefault}

%% Format Chapter headings
\usepackage{titlesec}
\titleformat{\chapter}[hang]{\normalfont\huge\bfseries\color{tcd_blue}}{\thechapter}{1cm}{}{}

%% Listings style for code and SPARQL
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny,
  tabsize=2
}

\title{\thesistitle}
\author{\authorname}

\hypersetup{
   pdftitle=\thesistitle,
   pdfauthor=\authorname,
   pdfkeywords=\keywords,
   pdfsubject=\degree,
   pdfinfo={
     pdfsupervisor=\supervisor,
   }
}

\frontmatter
\begin{document}
\input{title/title.tex}
\pagenumbering{roman}

%----------------------------------------------------------------------------------------
%	DECLARATION
%----------------------------------------------------------------------------------------
\section*{\Huge\textcolor{tcd_blue}{Declaration}}
\vspace{1cm}
I hereby declare that this \typeofthesis\ is entirely my own work and that it has not been submitted as an exercise for a degree at this or any other university.

\vspace{1cm}
I have read and I understand the plagiarism provisions in the General Regulations of the University Calendar for the current year, found at \url{http://www.tcd.ie/calendar}.
\vspace{1cm}

I have completed the Online Tutorial on avoiding plagiarism `Ready Steady Write', located at \url{http://tcd-ie.libguides.com/plagiarism/ready-steady-write}.
\vspace{1cm}

I consent to the examiner retaining a copy of the thesis beyond the examining period, should they so wish (EU GDPR May 2018).
\vspace{1cm}

I agree that this thesis will not be publicly available, but will be available to TCD staff and students in the University's open access institutional repository on the Trinity domain only, subject to Irish Copyright Legislation and Trinity College Library conditions of use and acknowledgement.
\vspace{3cm}

Signed:~\rule{5cm}{0.3pt}\hfill Date:~\rule{5cm}{0.3pt}

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------
\chapter*{Abstract}

The proliferation of AI systems across high-stakes domains has intensified pressure on research institutions, funding bodies, and developers to demonstrate meaningful ethics compliance. Yet the landscape these actors must navigate is fragmented, multi-layered, and operationally opaque. Requirements drawn from institutional review frameworks, EU legislation, research funding conditions, and professional codes of conduct exist in parallel, overlap in complex ways, and collectively resist systematic application at the speed modern research pipelines demand.

This dissertation addresses that gap by designing, building, and evaluating a working AI ethics assessment engine: the AI-Assisted Ethics Assessment Framework (AIEF). Given a research proposal as free-text input, the system automatically produces structured compliance assessments across four regulatory regimes: the Trinity College Dublin Research Ethics and Approval Management System (REAMS), the EU AI Act, Horizon Europe, and the ACM/NeurIPS community guidelines.

The technical architecture rests on three interacting layers. First, an OWL ontology comprising 125 classes, 22 object properties, and 16 data properties harmonises requirements from all four frameworks, aligned with the W3C Data Privacy Vocabulary (DPV) and the Open Digital Rights Language (ODRL). Second, a knowledge graph populated with 207 governance requirements and 70 real-world AI incidents from the AIAAIC repository, connected through 342 mappings to EU Charter of Fundamental Rights articles, provides the structured evidence base. Third, a retrieval-augmented generation (RAG) pipeline retrieves framework-specific requirements and incident precedents via SPARQL queries, then grounds large language model outputs in this structured knowledge to produce traceable, rights-linked compliance assessments.

Evaluation across 25 proposals (20 synthetic, 5 real-world cases including COMPAS, Amazon Rekognition, and the Optum algorithm) demonstrates retrieval recall of 0.99 across all model configurations, confirming the knowledge graph's contribution is model-agnostic. Risk classification accuracy reaches 90\% with Llama~3.3~70B and 65\% with Llama~3.1~8B, establishing that generation quality scales with model capacity while the underlying knowledge engineering remains stable. An ablation study isolating LLM-only, KG-only, and full pipeline conditions confirms that the knowledge graph provides requirement traceability, incident grounding, and rights coverage that are absent from LLM-alone baselines.

The ontology is published at a persistent URI (\url{https://w3id.org/aief/}) and the complete system is available as open-source software. To the best of our knowledge, no existing tool combines free-text proposal input, multi-framework coverage, an OWL ontology with incident grounding, EU Charter rights mapping, risk classification, and requirement-level traceability within a single RAG-based architecture; AIEF's contribution is the integration of these components into an operational, evaluated system rather than any one component in isolation. The system realises the automated compliance assessment tool proposed as future work by Pandit and Rintam\"{a}ki (2024) and provides a foundation for extending automated ethics assessment to corporate AI governance and regulatory auditing contexts.


%----------------------------------------------------------------------------------------
%	LAY ABSTRACT
%----------------------------------------------------------------------------------------
\chapter*{Lay Abstract}

When researchers build artificial intelligence systems, they must demonstrate that their work is ethical and complies with a range of rules set by universities, governments, funding agencies, and professional organisations. At present, this process is almost entirely manual: a researcher must read through multiple lengthy policy documents, identify which rules apply to their specific project, and produce documentation showing they have considered the relevant ethical issues. This is time-consuming, error-prone, and often inconsistent across different projects.

This dissertation presents a software tool that automates this process. A researcher writes a plain-language description of their proposed AI project and submits it to the system. The tool then automatically identifies which ethics rules apply from four different sources of regulation, determines which fundamental human rights may be affected, retrieves examples of past AI systems that caused real-world harm in similar contexts, classifies the overall risk level, and produces a structured report listing specific compliance requirements with traceable references.

The system was tested on 25 research proposals and correctly identified the risk level 90\% of the time when using a sufficiently capable AI model. Critically, the structured knowledge base underpinning the tool performed consistently regardless of which AI model was used to generate the final assessment, demonstrating that the knowledge engineering is reliable and reusable. The tool, its underlying knowledge base, and its source code are all publicly available.


\newpage
\onehalfspacing\raggedright

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------
\section*{\Huge\textcolor{tcd_blue}{Acknowledgements}}

I would like to express my sincere gratitude to my supervisor, Prof. Dave Lewis, for his guidance and support throughout this research. His expertise in data governance and the ADAPT Centre's research environment provided the ideal context for this work.

I am particularly grateful to Delaram Golpayegani for her detailed ontology review, her seven-point feedback that substantially improved the knowledge representation, and her generosity in sharing her expertise on the FRIA ontology and AIRO. I also wish to thank Monique Munarini for her insights on fundamental rights assessment methodology and for introducing the synthetic participant validation approach that shaped the evaluation design.

I extend my thanks to Harsh Pandit, whose work on the FRIA ontology and Fundamental Rights Impact Assessment provided the intellectual foundation upon which this system builds.

Finally, I would like to thank my family and friends for their unwavering support throughout this programme.


\tableofcontents
\listoffigures
\listoftables

\newpage
\section*{\Huge\textcolor{tcd_blue}{Nomenclature}}
\begin{tabular}{lp{10cm}}
AIAAIC & AI, Algorithmic, and Automation Incidents and Controversies Repository \\
AIEF & AI-Assisted Ethics Assessment Framework \\
AIRO & AI Risk Ontology \\
DPV & Data Privacy Vocabulary \\
DSR & Design Science Research \\
FRIA & Fundamental Rights Impact Assessment \\
KG & Knowledge Graph \\
LLM & Large Language Model \\
ODRL & Open Digital Rights Language \\
OWL & Web Ontology Language \\
RAG & Retrieval-Augmented Generation \\
REAMS & Research Ethics and Approval Management System \\
SPARQL & SPARQL Protocol and RDF Query Language \\
TTL & Turtle (RDF serialisation format) \\
URI & Uniform Resource Identifier \\
\end{tabular}

\mainmatter
\input{introduction/introduction.tex}
\input{background/background.tex}
\input{project/project.tex}
\input{evaluation/evaluation.tex}
\input{conclusion/conclusion.tex}

\bibliographystyle{plain}
\bibliography{bibs/references}

\appendix
\renewcommand{\thechapter}{A\arabic{chapter}}
\input{appendix/appendix.tex}

\end{document}
